Formal Modelling of a Microcontroller Instruction Set in B

### 2009-07-01 trunk/papers/SBMF2009/source/paper.tex
Final submetido

The B method~\cite{Abrial} supports the construction of safety systems
model by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

Essentially, a B module contains two main parts: a header and the
available operations. Figure~\ref{fig:maqB} has a very basic
example. The clause $\mathit{MACHINE}$ has the name of module.  The
following two clauses respectively reference external modules and
create an instance of an external module. The $\mathit{VARIABLES}$
declares the name of the variables that compose the state of the
machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $
sepecifies the initial states. Finally, operations correspond to
the transitions between states of the machine.

@@

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural_vertical2.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

@@

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because it can set many
safety rules. However, the model can be very difficult to prove.

@@

However, few proof commands\footnote{The proof commands are steps that direct the
prover to find the proof, and cannot introduce false hypothesis.} need to be used
to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of proofs WD. We also set up a proving environment
consisting of networked computers to take advantage of the distribution
facilities now provided in the B development environment. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the virtual machines
using the B method. Then, the B method has been used successfully to model the
operational semantic. However the cost of modelling was still expensive and this
paper quoted some techniques to lower the cost of modelling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

@@

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus, some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not the micro-architecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.

@@

This work has shown an approach to the formal modelling of the
instruction set of microcontrollers using the B method. During the
construction of this model, some ambiguities and errors were
encountered in the official reference~\cite{Z80_manual}. As
the B notation has a syntax that is not too distant from that of
imperative programming languages, such model could be used to improve
the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the
correctness of typing, the well-definedness of expressions, in addition
to safety properties of the microcontroller state.

>>>>

The B method~\cite{Abrial} supports the construction of safety systems
models by verification of proofs that guarantees its correctness. So,
an initial abstract model of the system requirements is defined and
then it is refined until the implementation model. Development
environments based on the B method also include source code generators
for programming languages, but the result of this translation
cannot be compared by formal means. The paper \cite{Dantas_SBMF08} presented
recently an approach to extend the scope of the B method up to the
assembly level language. One key component of this approach
is to build, within the framework of the B method, formal models of
the instruction set of such assembly languages.

@@

Essentially, a B module contains two main parts: a header and the available
operations. Figure~\ref{fig:maqB} has a very basic example. The clause
$\mathit{MACHINE}$ has the name of module.  The next two clauses respectively
reference external modules and create an instance of an external module. The
$\mathit{VARIABLES}$ clauses declares the name of the variables that compose the
state of the machine. Next, the $\mathit{INVARIANT}$ clause defines the type and
other restrictions on the variables. The $\mathit{INITIALIZATION} $ sepecifies
the initial states. Finally, operations correspond to the transitions between
states of the machine.

@@

\begin{figure}[h] \centering
\includegraphics[width=.92\textwidth]{diagramaEstrutural.png}
 \caption{Dependency diagram of the Z80 model.}
\label{fig:hardware-definition-graph}
\end{figure}

@@

The proof obligations allow to verify the data types, important system
properties and if the expressions are well-defined (WD)\footnote{An
  expression is called ``well-defined'' (or unambiguous) if its
  definition assigns it a unique interpretation or value.}. The
properties provide additional guarantees, because they can set many
safety rules. However, the model can be very difficult to prove.

@@

However, few proof commands\footnote{The proof commands are steps that direct the
prover to find the proof, and cannot introduce false hypothesis.} need to be used
to prove most proof obligations. As there are many similar assembly instructions,
some human-directed proofs, when replayed, could discharge other proof
obligations. A good example is a set of 17 proof commands that quickly aided the
verification of 99\% (2295) of WD proofs. We also set up a proving environment
consisting of networked computers to take advantage of the distribution
facilities now provided in the B development environment. Finally, all of the
2926 proof obligations were proved using the tool support of the development
environment.

@@

There are in the literature of computer science some
approaches~\cite{BHDL_2003,GEMPLUS_99} to model hardware and the virtual machines
using the B method. Then, in both works the B method has been used successfully
to model the operational semantic. However the cost of modelling was still expensive and this
paper quoted some techniques to lower the cost of modelling.
% Although, the modelling of the Z80 do not have modelled some small aspects.

@@

The main motivation of our research is the development of verified
software up to the assembly level, which requires specifying the
semantics of the underlying hardware. Thus, some aspects were not
modelled in our work such as the execution time of the instructions.
Also we did not consider the microarchitecture of the hardware as the scope of
our work does not include hardware verification. However, there are
many other specialized techniques to verify these questions.

@@

This work has shown an approach to the formal modelling of the instruction set of
microcontrollers using the B method. During the construction of this model, some
ambiguities and errors were encountered in the official reference for Z80
microcontroller~\cite{Z80_manual}. As the B notation has a syntax that is not too
distant from that of imperative programming languages, such model could be used
to improve the documentation used by assembler programmers. Besides, the formal
notation used is analyzed by software that guarantees the correctness of typing,
the well-definedness of expressions, in addition to safety properties of the
microcontroller state.
